---
title: "PM566LAB3Sabrina"
format: html
editor: visual
embed-resources: true
theme: darkly
---

## Step 1 Read in the data & Step 2 Check the dimensions, headers, footers

```{r}
met <- read.csv("D:/USC-MPH2025/PM566/PM566Lab/met_all.gz")

dim(met)
head(met)
tail(met)
```

## Step 3 Take a look at the variables

```{r}
str(met)
```

## Step 4 Take a closer look at the key variables

```{r}
table(met$year)
table(met$day)
table(met$hour)
summary(met$temp)
summary(met$elev)
summary(met$wind.sp)
```

```{r}
met$elev[met$elev==9999.0] <- NA
summary(met$elev)
```

```{r}
met <- met[met$temp > -40, ]
met2 <- met[order(met$temp), ]
head(met2)
```

## Step 5 Check the data against an external data source.

```{r}
met <- met[met$temp > -15, ]
met2 <- met[order(met$temp), ]
head(met2)
```

## Step 6 Calculate summary statistics

```{r}
elev <- met[met$elev == max(met$elev, na.rm = TRUE),]
summary(elev)
```

```{r}
cor(elev$temp, elev$wind.sp, use="complete")
cor(elev$temp, elev$hour, use="complete")
cor(elev$wind.sp, elev$day, use="complete")
cor(elev$wind.sp, elev$hour, use="complete")
cor(elev$temp, elev$day, use="complete")
```

## Step 7 Exploratory graphs

```{r}
hist(met$elev, breaks=100)
hist(met$temp)
hist(met$wind.sp)
```

```{r}
library(leaflet)
leaflet(elev) %>%
  addProviderTiles('OpenStreetMap') %>% 
  addCircles(lat=~lat,lng=~lon, opacity=1, fillOpacity=1, radius=100)
```

```{r}
library(lubridate)
elev$date <- with(elev, ymd_h(paste(year, month, day, hour, sep= ' ')))
summary(elev$date)
elev <- elev[order(elev$date), ]
head(elev)
```

```{r}
plot(elev$date, elev$temp, type='l')
plot(elev$date, elev$wind.sp, type='l')
```

Temperature at the station with maximum elevation shows a clear daily cycle, where repeated rises during daytime and falls at night, and stayed relatively stable conditions across the month of August, with no strong overall trendings.\
Wind speed shows a rough daily cycle but is more unstable than temperature, with visible short-term fluctuations rather than a smooth pattern, such as notably sharp plunges around August 19 and August 26, and a gradual increase in wind speed between August 12 and August 18. No consistent trend is apparent over the full month beyond these localized swings.

## Step 8 Ask questions

1.  What differences exist between the temperature or wind speed patterns at this highest elevation station and those at the lowest elevation station?
2.  Temperature shows a moderate correlation with hour (r=0.44), but wind speed shows almost no correlation with hour (r = 0.088). Why doesn't wind speed follow a stable daily cycle the way temperature does at this highest elevation station? If the situation different in other stations or other months.
